\vspace{1.0cm} \LARGE \bf The ability of single genes vs full genomes to resolve time and space in outbreak analysis

### 2019-08-05 revisions_1/response.tex

2.1. ``Related to the previous point, I find it troubling that the MCMC chains apparently did not converge. I am not saying that the analyses are invalid, but with what is written in the manuscript I have to conclude that chains did not mix and that convergence was not reached. Please give some clarification about which parameters did not mix. I wouldn't expect the tip-dates to mix, but I would expect all of the other parameters to. If parameters are not mixing this is either because analyses were not run long enough (and should be run longer), operators are not making good proposals (and should be tweaked/changed) or because the model is not identifiable (see Bruce Rannala's 2002 paper about identifiability in Systematic Biology).''

\textbf{We apologise for very imprecise language in the manuscript. Indeed, the issues of convergence only affect tip date inference, whereas other parameters suffer from bad mixing. We have clarified this throughout the manuscript.}

2.2. ``I note that this is especially worrying to me because the lead author published a paper analysing the full 1,610 EBOV sequence dataset under roughly the same model (the difference here is that some tip-dates and -locations are unknown). I couldn't find any notes/statistics about convergence in that paper, and without further explanation the results reported here calls into question the results of the previous paper. (Clarify if the non-convergence here is due to differences in the model used or if the original analysis also did not converge).''

\textbf{Not mentioning convergence/mixing in the previous paper was an error on our part and we apologise for that. There were a number of factors which helped mixing in our previous paper, such as full information (\textit{i.e.} 10\% of data are not missing), having more collection dates to restrict parameter space, and better hardware to run analyses within reasonable time. Even then some parameters (likelihood for intergenic regions and random effects) had ESSs above 100 but below 200 with likelihood for first codon position having the worst ESS at 43. For that particular paper we have run numerous models with the same data over the course of the study and the mixing issues had always been there. Having said, despite numerous iterative refinements we have never seen MCMC run for the 2017 paper spuriously converge on anything that would change our conclusions.}

@@

3. ``How did you compute the HPDs for multi-peaked posterior distributions? It is mentioned that a custom Python script was used, but not what this script actually does. It sounds like the posteriors are multi-peaked because you combined chains that converged to different peaks, not because all chains were mixing across all of the different peaks, so the authors' strategy here needs to be explained in more detail.''

\textbf{The posteriors for tip dates had examples of both chains sampling from two to three distributions independently and chains where each run converged onto one of two or three distributions. We now explain this in the text and provide a description for how the disjoint posteriors were computed. Briefly, it involves an iterative approach where we intersect peaks of kernel density estimates, compute the integral of areas intersected and broaden the search until the peaks intersected capture 0.95 of the area under the kernel density estimate.}

@@

9. ``Without an x-axis or grid lines of some sort the right-hand panels in Fig. 3 are not informative. At present only tips where the 95\% HPD contains the true location are highlighted. Why not also show those tips where the most probable inferred location is the true location? (as that is the metric actually used in the comparison).''

\textbf{Agreed, apologies for that. We have altered Fig. 3 to encode information both about the 95\% most probable set and whether the best model guess is the truth in a seamless way. We have also added grid lines every 0.2 to the right-hand panels.}

10. ``Isn't it possible to summarise all of the statistics computed on the estimated tip-dates and -locations in a table or boxplots?''

\textbf{We have included an extra supplementary figure (Fig S8) that depicts all of the major statistics reported in the manuscript.}

@@

12. ``HIV-1 is conspiciously absent from Fig. 6. As is pointed out in the introduction it is still very common to only sequence the part of pol with resistance mutations. Older studies tended to sequence only env or gag and full-length HIV genomes still make up only a small amount of the sequences in public databases. The same is true for HCV, where the core/E1 and NS5B genes are commonly sequenced, but only the full genome is shown in Fig. 6.''

\textbf{Yes, thank you. This is entirely an oversight on our part, because HIV was meant to be included from the start. We have now included pol, env and coding regions of HIV and NS5B region of HCV into the figure. We have refrained from adding any more on account of the plot already getting a bit crowded.}

13. ``There has been a fair amount of interest in doing phylodynamics with bacterial genomes, especially tuberculosis, and the argument is often made that because the genomes are so much longer it is not necessary to only look at very long timescales. I did a quick back-of-the-envelope calculation and got ~3.5 years for the temporal horizon of TB (I may have made a mistake, and I considered the genome length, not just variable sites). This obviously won't fit into the figure, but it would be nice to have a note about the temporal horizon for bacterial genomes in the text.''

\textbf{We agree that bacteria have joined the ranks of measurably evolving populations thanks to genomics. The estimated temporal horizon for TB the reviewer has calculated is approximately correct - quickly googling for parameters of interest (genome length = 4.4Mb, evolutionary rate = $1\times10^{-7}$) it's 2.27 years (1/(genome length * evolutionary rate)). We have included a section in the discussion that mentions a couple of bacterial examples, though, as we note, rate estimates in bacteria can have a wide margin of error depending on how they were calculated, and, as it is clear from the cited M. tuberculosis and what we have reported in this response, even reported genome lengths can differ meaningfully.}

@@

\textbf{Correct, thank you for spotting the mistake.}

%%% REVIEWER 2 %%%
\section*{Reviewer 2}

1. ``Equation 2 is not really a rearrangement of Eq. 1 because the P term is missing. This only requires a short line to explain that Eq. 2 assumes that there has been at least one mutation.''

\textbf{Yes, we have indeed missed a step in our logic. We have added the missing step and explained it in the text. The missing step was taking the mean expectation, so that an arbitrary threshold for the probability of observing a mutation would not need to be established. The mean of an expontential distribution is just 1/rate.}

@@

4. ``In lines 58 to 62, page 3 and throughout the rest of the manuscript there is some discussion about how Bayesian inference can resolve branching times, but maximum likelihood cannot. This is not necessarily the case, BEAST does use a tree prior that is rooted and based on a demographic/branching process, which means that all trees are chronograms, mostly with non-zero branch lengths. MrBayes can use a flat tree prior on topologies and an unconstrained model, so the trees are effectively 'phylograms' (branch lengths in subs/site). Therefore, the resolution of branching times is not inherent to Bayesian methods, it is just the result of the branching model (i.e. tree prior). Moreover, a molecular clock method in maximum likelihood, like LSD would still fail to resolve these branches. If one took a consensus tree from BEAST, it is likely that these branches would be collapsed into a polytomy, as it seems to be the case in maximum likelihood.''

\textbf{Yes, we agree too and apologise for the misunderstanding. We have altered the text to reflect that resolving branching order is performed via the tree prior rather than as a result of Bayesian inference.}

5. ``Line 21, page 8 states "sequences are assigned to compartments that are too small", what do the authors mean by 'compartments'?''

\textbf{We are sorry for the confusion - `compartments' in that sentence meant states, \textit{i.e.} if more states are introduced into a model the finer granularity of the model might not be matched by the temporal resolution available in the data. We have added this explanation to the text.}

### 2019-08-05 genomic-horizon.tex
Finalise revision 1.

Molecular clocks have been particularly useful in molecular epidemiology, where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
% Mutations, prior to the action of selection, occur randomly across a genome via misincorporation of nucleotides during replication by the polymerase.
% Many of these mutations are filtered out by selection if they adversely affect protein function, with leftover mutations being beneficial, neutral, or mildly deleterious.
Generally, neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics, unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature, mutations are modelled as a Poisson process \citep{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently, the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Since the probability of observing at least one mutation changes depending on waiting time (and \textit{vice versa}) we instead focus on the mean waiting time until at least one mutation appears.
The mean of an exponential distribution is $\lambda^{-1}$ where $\lambda$ in our case is \textit{RL}, such that mean waiting time $\bar{t}$ under a given evolutionary rate \textit{R}, and sequence length \textit{L} becomes
\begin{equation}
  \bar{t} = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times $\bar{t}$ are lengthened and \textit{vice versa}.
It also suggests a worrying relationship between \textit{R}, \textit{L}, and $\bar{t}$ - a reduction in either \textit{R} or \textit{L} leads to a reciprocal reduction in $\bar{t}$, such that reducing the number of sites by 10\%, for example, requires a $\frac{1}{0.9} = 1.11(1)$-fold increase in evolutionary rate to maintain the same temporal resolution $\bar{t}$ or risk increasing the waiting time for a mutation by the same amount (\textit{i.e.} $\approx$11\%, or 0.9 of the original resolution available).
This gets worse reciprocally, such that halving the sites requires doubling the evolutionary rate, using 10\% of the sites requires a ten-fold increase in evolutionary rate, \textit{etc}.

@@

% Because of these natural constraints there is thus a lower limit to how many mutations can be expected to occur in a given length of sequence evolving at a particular evolutionary rate.
% We thus refer to the mean waiting time for a mutation as the temporal horizon, which reflects a theoretical temporal resolution over which mutations arise and are available for phylogenetic inference in a study system.

In this study we show this by quantifying how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of >1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \citep{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split, where we re-infer these parameters as latent variables using Markov chain Monte Carlo (MCMC).
We show that this reduction in data not only leads to severe mixing issues in MCMC analyses by removing the constraints additional data impose on plausible parameter space without adding restrictive priors to compensate, but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity, even with genome sequences.
% As such, pathogen genomes have an inherent temporal resolution limit, where processes occurring at a rate lower than the rate of evolution of a given pathogen will not be recovered reliably using currently standard phylogenetic methods.
% We refer to this as the temporal horizon and express it as the mean waiting time until at least one mutation occurs and differentiates a pair of evolving sequences.
 % which is already substantial for fine scale epidemiology and that sequencing subgenomic fragments of a pathogen results in loss of information that can make sequence data unreliable beyond diagnostics and typing.
% In order to highlight how much information is lost and to relate the concept of temporal horizon with real data we use a subset of a well-characterised dataset of Ebola virus genomes from West Africa collected in 2014-2015 \citep{dudas_virus_2017}.
% By analysing both complete viral genomes and the same dataset reduced to just the surface glycoprotein GP (corresponding with just 10\% of all genomic sites), we show substantial loss of predictive power, incorrect or unreliable reconstructions of past events, and migration model, despite increased evolutionary rate in GP compared to complete genome.
% We hope that analyses presented here may serve as a guide to researchers in the field of genetic epidemiology of how to think about the statistical power of sequence data itself, as well as encouragement to model developers to address the widening rift between data volumes and methods used to analyse them.

@@

Similarly, locations are inferred correctly more often with complete genomes than with GP sequences, where the maximum probability location (\textit{i.e.} the model's best guess) matches the truth, where using complete genomes results in 0.540 probability of guessing correctly compared to 0.286 probability for GP (for a calibration of both models see Figure \ref{calibration}).
The model makes these guesses with more certainty too, where the mean probability of the true location is 0.482 with genomes and 0.219 with GP, and mean probability of best guess (\textit{i.e.} maximum probability) is 0.680 and 0.396, respectively.
% \tbc{Need to more carefully define what you mean by `inferred correctly'. Cross-entropy does this well, otherwise need a cut-off for assignment. Over 50\%?}
We also calculated the great circle distances between the population centroids of true and each predicted location weighed by probability, which should ideally be 0.0 (0 km distance multiplied by probability of 1.0).
The mean of these distances across masked tips are 75.886 kilometres for genomes compared to 164.309 km for GP sequences.

@@

% One of these is strain `14859\_EMLK', a virus descended from the initial sweep of Ebola virus through Sierra Leone via Kailahun and Kenema districts which eventually ended up in Conakry (Guinea) from where it jumped back into Sierra Leone late in the epidemic where it was sequenced \citep{arias_rapid_2016}.
% `EM\_004422', which migrated through Kailahun district of Sierra Leone too, but then spread into Liberia from which it spilled back into Guinea via Macenta prefecture before ending up in Kissidougou (Guinea) where it was sequenced \citep{carroll_temporal_2015}.
% Similar to other lineages, `MK3462' was part of the sweep through Sierra Leone, though unlike others remained in-country, migrating to the environs of Freetown and then Bombali district where it was sequenced \citep{arias_rapid_2016}.
% `PL5294', unlike others, belonged to a lineage that was not part of the Sierra Leonean sweep and instead part of an unusual and under-sampled lineage endemic to western Guinea (``lineage A'', \citep{carroll_temporal_2015}), from where it spilled into Sierra Leone's Kambia district late in the epidemic \citep{arias_rapid_2016}.

@@

\subsubsection*{Bayesian analyses}
Both GP and genome datasets were analysed in BEAST v1.8.10 \citep{suchard_bayesian_2018} under the generalised linear model (GLM) described previously \citep{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017} to infer the migration model.
Sites in both GP and genome alignments were partitioned into codon positions 1, 2, and 3, with the genome analysis also including a partition comprised of non-coding intergenic regions.
Each partition was assigned an independent HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} substitution model.
A relaxed molecular clock \citep{drummond_2006} with an uninformative prior on the mean \citep{ferreira_bayesian_nodate} of the log-normal distribution was used as the clock model.
A flexible skygrid tree prior \citep{gill_2013} was used to infer estimates of effective population size across 100 evenly spaced points in time starting 1.5 years prior to the collection of the most recent sequence to the date of the most recent sequence.

@@

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{supp_figures/sfigX_treetimeLocations.png}
	\caption{\textbf{Maximum likelihood inference of masked sequence location from genomes (left) and GP sequences (right) via a CTMC model implemented in TreeTime.}
  Unlike the Bayesian GLM where probabilities are spread across different locations, the CTMC gives categorical support to one location.
  Genomes still perform better in terms of correct guess (0.432 probability that best guess location is true location for genomes versus 0.259 for GP), cross entropy (12012.800 nats for genome versus 24397.109 nats for GP) and mean probability-weighted great circle distance between true location population centroid and estimated location population centroid (87.568 km for genome versus 124.909 km for GP).
	}
	\label{TTlocations}
\end{figure}

>>>>

Molecular clocks have been particularly useful in molecular epidemiology, where the accumulation of mutations between sequences is used as a noisy approximation for elapsed time, given either times of events in the phylogeny (sequence dates or dates of common ancestors) or a previously determined molecular clock rate.
Generally, neutral pathogen variation at the nucleotide level ebbs and flows under the forces of population genetics, unlike beneficial or deleterious variation which tends to either fix or be purged rapidly, respectively.
Due to their random and discrete nature, mutations are modelled as a Poisson process \citep{yang_computational_2006} where the waiting time $t$ for observing a mutation at a single site is exponentially distributed with evolutionary rate parameter $R$.
The probability of observing 0 mutations at a single site after time $t$ is $e^{-Rt}$ and the probability of at least one mutation is therefore $1-e^{-Rt}$.
Higher evolutionary rates \textit{R} or waiting times \textit{t} result in higher probabilities of observing at least one mutation at the site in question.
Since sites are assumed to evolve independently, the probability of observing at least one mutation across \textit{L} sites is
\begin{equation}
  P = 1-e^{-RLt},
\end{equation}
where \textit{RL} is expressed in substitutions per year (rate in substitutions per site per year multiplied by number of sites).
Since the probability of observing at least one mutation changes depending on waiting time (and \textit{vice versa}) we instead focus on the mean waiting time until at least one mutation appears.
The mean of an exponential distribution is $\lambda^{-1}$ where $\lambda$ in our case is \textit{RL}, such that mean waiting time $\bar{t}$ under a given evolutionary rate \textit{R}, and sequence length \textit{L} becomes
\begin{equation}
  \bar{t} = \frac{1}{RL}
  \label{horizon}
\end{equation}
When the evolutionary rate \textit{R} or sequence length \textit{L} are low mean waiting times $\bar{t}$ are lengthened and \textit{vice versa}.
It also suggests a worrying relationship between \textit{R}, \textit{L}, and $\bar{t}$ - a reduction in either \textit{R} or \textit{L} leads to a reciprocal reduction in $\bar{t}$, such that reducing the number of sites by 10\%, for example, requires a $\frac{1}{0.9} = 1.11(1)$-fold increase in evolutionary rate to maintain the same temporal resolution $\bar{t}$ or risk increasing the waiting time for a mutation by the same amount (\textit{i.e.} $\approx$11\%, or 0.9 of the original resolution available).
This gets worse reciprocally, such that halving the sites requires doubling the evolutionary rate, using 10\% of the sites requires a ten-fold increase in evolutionary rate, \textit{etc}.

@@

In this study we show this by quantifying how much information relevant to phylodynamic analysis is lost when shorter genomic regions are used instead of full genomes.
By focusing our attention on a subset (600 sequences) of a well-characterised genomic sequence data (comprised of >1600 viral genomes) set derived from the West African Ebola virus epidemic of 2013-2015 \citep{dudas_virus_2017} we estimate loss in precision and accuracy of molecular clock models and phylogeographic inference methods when only the glycoprotein gene (GP), a region representing just 10\% of the viral genome, is analysed despite GP evolving at rates faster than the genomic average.
Our methods rely on masking tip dates and locations for 60 (10\%) of the sequences in a classic training-testing split, where we re-infer these parameters as latent variables using Markov chain Monte Carlo (MCMC).
We show that this reduction in data not only leads to severe mixing issues in MCMC analyses by removing the constraints additional data impose on plausible parameter space without adding restrictive priors to compensate, but can also result in unreliable tip date and location inference.
Despite achieving much better temporal resolution when using complete viral genomes we still find residual error caused by inherent randomness of mutations which is close to theoretical expectations (Eq \ref{horizon}).
We refer to this as the temporal horizon, \textit{i.e.} a temporal resolution limit where population processes occurring at a rate faster than the rate at which mutations enter and are observed in a population will not be captured with high fidelity, even with genome sequences.

@@

Similarly, locations are inferred correctly more often with complete genomes than with GP sequences, where the maximum probability location (\textit{i.e.} the model's best guess) matches the truth, where using complete genomes results in 0.540 probability of guessing correctly compared to 0.286 probability for GP (for a calibration of both models see Figure \ref{calibration}).
The model makes these guesses with more certainty too, where the mean probability of the true location is 0.482 with genomes and 0.219 with GP, and mean probability of best guess (\textit{i.e.} maximum probability) is 0.680 and 0.396, respectively.
We also calculated the great circle distances between the population centroids of true and each predicted location weighed by probability, which should ideally be 0.0 (0 km distance multiplied by probability of 1.0).
The mean of these distances across masked tips are 75.886 kilometres for genomes compared to 164.309 km for GP sequences.

@@

@@

\subsubsection*{Bayesian analyses}
Both GP and genome datasets were analysed in BEAST v1.8.10 \citep{suchard_bayesian_2018} under the generalised linear model (GLM) described previously \citep{faria_simultaneously_2013,lemey_unifying_2014,dudas_virus_2017} to infer the migration model.
Sites in both GP and genome alignments were partitioned into codon positions 1, 2, and 3, with the genome analysis also including a partition comprised of non-coding intergenic regions.
Each partition was assigned an independent HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} substitution model.
A relaxed molecular clock \citep{drummond_2006} with an uninformative CTMC reference prior on the mean \citep{ferreira_bayesian_nodate} of the log-normal distribution was used as the clock model.
A flexible skygrid tree prior \citep{gill_2013} was used to infer estimates of effective population size across 100 evenly spaced points in time starting 1.5 years prior to the collection of the most recent sequence to the date of the most recent sequence.

@@

\begin{figure}[h]
 \centering
	\includegraphics[width=0.75\textwidth]{supp_figures/sfigX_treetimeLocations.png}
	\caption{\textbf{Maximum likelihood inference of masked sequence location from genomes (left) and GP sequences (right) via a CTMC model implemented in TreeTime.}
  Horizontal bars indicate the posterior distribution of masked tip locations, coloured by country (Sierra Leone in blue, Liberia in red, Guinea in green) and location (lighter colours indicate administrative divisions lying towards west of the country).
  The correct location of each tip is outlined in white, with the smaller plot to the right showing only the probability of the correct location.
  Bars marked with an open circle indicate cases where the correct location is within the 95\% credible set and solid circles indicate cases where the location with the most probability is also the correct location.
  Genomes still perform better in terms of correct guess (0.432 probability that best guess location is true location for genomes versus 0.259 for GP), cross entropy (12012.800 nats for genome versus 24397.109 nats for GP) and mean probability-weighted great circle distance between true location population centroid and estimated location population centroid (87.568 km for genome versus 124.909 km for GP).
	}
	\label{TTlocations}
\end{figure}
